---
title: "DATA 643 Project 1 | Global Baseline Predictors and RMSE"
author: "Jason Joseph, Srini Illapani"
date: "June 11, 2017"
output: 
  html_document:
    fig_caption: yes
    highlight: pygments
    theme: cerulean
    toc: yes
---

```{r message=FALSE, warning=FALSE}

library(pander)
library(ggplot2)
library(knitr)
library(dplyr)
```

### 1.0 Build a basic recommender system

This is a basic recommender system where we created a sample ratings grid where few books have been rated by discrete users.



### 2.0 Load the data & Analyse

Load the sample data set with ratings for books and convert the same into a matrix format.



```{r message=FALSE, warning=FALSE}

# load csv into data variable 
# https://raw.githubusercontent.com/isrini/SI_DATA643/master/book_ratings.csv
data <- read.csv("book_ratings.csv",row.names = 1)

# convert data into a matrix
data <- as.matrix(data)
pander(data)

```


#### 2.1 Top 3 Books

Here are the top 3 rated books


```{r message=FALSE, warning=FALSE}

means <- colMeans(data, na.rm = TRUE)
cols <- colnames(data)[order(means, decreasing = TRUE)[1:3]]
top3 <- data.frame(books = cols,stringsAsFactors = FALSE)
pander(top3)

```



### 3.0 Data Split

Lets split the data set in to two. Training and Test.
we selected 12 reviews from training. we will replace those with NA in the training set. NA was used so it would be omitted from our calculations. In the test dataset we only kept values identified for testing. the others were replaced with NA.


```{r}

test_rows <- c(1,3,4,5,6,7,14,13,19,20,12,14)
test_cols <- c(1,4,2,3,4,5,2,3,3,4,5,3)
test_indices <- cbind(test_rows,test_cols)

data_train <- data
data_train[test_indices] <- NA

data_test <- data
data_test[test_indices] <- 0
data_test[data_test > 0] <- NA
data_test[test_indices] <- data[test_indices]
```

#### 3.1 TRAIN DATA

```{r}
data_train
```


#### 3.2 TEST DATA

```{r}
data_test
```


### 4.0 Calculations


Find the mean, bias and RMSE for both the data sets


#### 4.1 Mean ratings of each User(Reader) with chart for Train data set


```{r message=FALSE, echo=FALSE, warning=FALSE}
user_means <- rowMeans(data_train,na.rm = TRUE)
user_means_df <-  data.frame(as.list(user_means))

# change user means from wide to long 
user_means_df <- tidyr::gather(user_means_df,"user") 
p1 <- ggplot(user_means_df,aes(x=user, y=value,fill=user))+ geom_bar(stat="identity") + labs(title="Plot of Mean User ratings",x="User",y="Avg. Rating")

colnames(user_means_df) <-c("User","Rating")
pander(user_means)

p1

```

#### 4.2 Mean ratings of each book with chart for Train data set


```{r message=FALSE, echo=FALSE, warning=FALSE}

book_means <- colMeans(data_train,na.rm = TRUE)
book_means_df <-  data.frame(as.list(book_means))

# change user means from wide to long 
book_means_df <- tidyr::gather(book_means_df,"book") 
p2 <- ggplot(book_means_df,aes(x=book, y=value,fill=book))+ geom_bar(stat="identity") + labs(title="Plot of Book Average Rating",x="Book",y="Avg. Rating")

colnames(book_means_df) <-c("Book","Rating")
pander(book_means)

p2
```


#### 4.3 Raw Averages

Rating for every user-item combination. For Test and Train data sets


```{r message=FALSE, echo=FALSE, warning=FALSE}
raw_test <- mean(data_test,na.rm = TRUE)
raw_test_mat <- data_test
raw_test_mat[] <- raw_test
raw_test

raw_train_mat <- data_train 
raw_train <- mean(data_train,na.rm = TRUE)
raw_train_mat[] <-raw_train
raw_train
```

#### 4.4 RMSE for raw averages

For Test and Train data sets


```{r message=FALSE, echo=FALSE, warning=FALSE}
#find squre difference 
squareDiff <- (data_train - raw_train_mat)^2

# find mean of squareDiff
squareDiff_mean <- mean(squareDiff,na.rm = TRUE)

#find square root
rmse_train <- sqrt(squareDiff_mean)

# train test 

squareDiff_test <- (data_test - raw_test_mat)^2
# find mean of squareDiff
squareDiff_test_mean <- mean(squareDiff_test,na.rm = TRUE)
#find square root
rmse_test <- sqrt(squareDiff_test_mean)

```

#### 4.5 Train Data

RMSE for Train

```{r message=FALSE, echo=FALSE, warning=FALSE}
rmse_train
```

#### 4.6 Test Data

RMSE for Test

```{r message=FALSE, echo=FALSE, warning=FALSE}
rmse_test
```


####  4.7 Bias for each user and book

User Bias

```{r}
## user bias
user_bias <- user_means - raw_train
user_bias_df <-  data.frame(as.list(user_bias))
user_bias_df <- tidyr::gather(user_bias_df,"user")

colnames(user_bias_df) <-c("User","Bias")
pander(user_bias_df)

```


Book Bias

```{r}
#book bias
book_bias <- book_means - raw_train
book_bias_df <-  data.frame(as.list(book_bias))

book_bias_df <- tidyr::gather(book_bias_df,"book")
colnames(book_bias_df) <-c("Book","Bias")
pander(book_bias_df)

```



### 5.0 Baseline predictors for each user item


```{r}

# raw average + user bias + book bias
calBaseLine <- function(in_matrix, book_bias_in,user_bias_in,raw_average)
{
  out_matrix <- in_matrix
  row_count <-1
  for(item in 1:nrow(in_matrix))
  {
    col_count <-1
    for(colItem in 1: ncol(in_matrix))
    {
      #out_matrix[row_count,col_count] <- 0
     out_matrix[row_count,col_count] <- raw_average[1] + user_bias_in[[row_count]] +  book_bias_in[[col_count]]
      col_count <- col_count +1  
    }
    row_count <- row_count +1
  }
return(out_matrix)
}

base_pred <- calBaseLine(data_train,book_bias,user_bias,raw_train)
pander(base_pred)
```


### 6.0 RMSE for the baseline predictors for both training data and test data sets


```{r}

## test data

# finding Error
data_err <- data_test - base_pred
# squaring error
data_err <- (data_err)^2
#finding average 
data_rmse_test<- mean(data_err[test_indices])
# square root 
data_rmse_test<- sqrt(data_rmse_test)


## training data

# finding Error
data_err_train <- data_train - base_pred
# squaring error
data_err_train <- (data_err_train)^2
#finding average 
data_rmse_train <- mean(data_err_train,na.rm = TRUE)
# square root 
data_rmse_train<- sqrt(data_rmse_train)

```

#### 6.1 RMSE - TEST DATA

```{r}
data_rmse_test
```

#### 6.2 RMSE - TRAIN DATA

```{r}

data_rmse_train
```

### 7.0 Summary

Lets calculate the percentage improvements based on the original (simple average) and baseline predictor (including bias) RMSE numbers for both Test and Train data sets.

The results show that we see a 50% improvement in making a prediction for the ratings in the Training data set. Where as we see only 38% improvement in prediction for the Test data set. Both are positive hoewver the Training data set yielded better prediction.


```{r}

# Train data set

R1 <- rmse_train
Rb1 <- data_rmse_train

Per_Improv_Train <- (1-(Rb1/R1))*100
Per_Improv_Train

# Test data set

R2 <- rmse_test
Rb2 <- data_rmse_test

Per_Improv_Test <- (1-(Rb2/R2))*100
Per_Improv_Test

```









